---
title: "MA344-SP-2411779"
author: "Khushal"
date: "2025-04-20"
output: pdf_document
latex_engine: xelatex



---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE,Warning = FALSE)
```

```{r load-data, echo=FALSE}
# load data
data <- read.csv("MA334-SP-7_2411779.csv")
```

## [Question 1 = Data exploration]{.underline}

#### 1.1

```{r, echo=FALSE}


# find observations & variables
library(knitr)
kable(head(data, 5))

cat("Number of observations:", nrow(data), "\n")
cat("Number of variables:", ncol(data), "\n")
```

**Data Types**

-   **Integer** **:** age, educ , gender ,hrswork ,insure , metro ,nchild , union, marital
-   **Numeric :** wage
-   **categorical :** race , region

#### 1.2 Descriptive Statistics

```{r echo=FALSE}
# Load required package
library(knitr)

# Keep only numeric columns
numeric_data <- data[sapply(data, is.numeric)]

# Calculate descriptive statistics
summary_stats <- data.frame(
  Variable = names(numeric_data),
  Mean = round(sapply(numeric_data, mean, na.rm = TRUE), 2),
  Median = round(sapply(numeric_data, median, na.rm = TRUE), 2),
  Min = round(sapply(numeric_data, min, na.rm = TRUE), 2),
  Max = round(sapply(numeric_data, max, na.rm = TRUE), 2),
  SD = round(sapply(numeric_data, sd, na.rm = TRUE), 2)
)

# Show summary table
kable(summary_stats, caption = "Summary Statistics of Numeric Variables")

```

- The workforce encompasses individuals from 18 to 85 years with a median age of 40 which shows a sample with a primarily young to middle-aged demographic.

- There is substantial variation (SD = 1.5) in education levels which average to 1.8 years.

- The wage distribution reveals a right-skew pattern where the mean exceeds the median and demonstrates high variability with a standard deviation of 14.3 indicating income disparity.

- The dataset includes 60% male participants and 40% female participants where male participants are represented by 0 and female participants by 1.

- Health insurance coverage exists for 80% of the survey respondents.

- The Midwest representation stands at 35% which exceeds its proportion compared to other regions.

- The percentage of workers who belong to a union stands at 15%.

### 1.3 Distribution Visualizations (Base R)

```{r, fig.width=3, fig.height=3,echo=FALSE}
hist(data$wage, 
     prob = TRUE, 
     main = "Distribution of Hourly Wages",
     xlab = "Wage ($/hour)", 
     col = "skyblue", 
     border = "black")
lines(density(data$wage, na.rm = TRUE), col = "red", lwd = 2)
```

-   **Wage Distribution:**

Highly right-skewed with most wages clustered below \$30/hour

Several extreme outliers above \$70/hour (likely executives/specialists)

Median wage appears around \$18/hour (from boxplot)

```{r, fig.fig.width=3, fig.height=3 ,echo=FALSE}
plot(data$age, data$wage, 
     main = "Wage vs Age", 
     xlab = "Age", 
     ylab = "Wage ($/hour)", 
     col = "blue", pch = 19)
abline(lm(wage ~ age, data = data), col = "red", lwd = 2)
```

-   **Age Distribution:**

The workforce shows a roughly normal distribution with slight right skew

Peak workforce age is between 35-45 years

Notable presence of workers beyond typical retirement age (65+)

#### 1.4

```{r echo=FALSE,fig.width=6, fig.height=4}

# Select numeric columns
numeric_data <- data[sapply(data, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(numeric_data, use = "complete.obs")


install.packages("corrplot")
library(corrplot)

corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")
```

-   **Strongest Positive Correlations:**

Education-Wage (r = 0.35): Higher education levels correlate with higher wages

Age-Wage (r = 0.25): Older workers tend to earn more, though the relationship is moderate

-   **Notable Negative Correlations:**

Age-Hours Worked (r = -0.18): Younger workers tend to work more hours

Number of Children-Hours Worked (r = -0.12): Workers with more children work slightly fewer hours

-   **Non-Significant Relationships:**

Education level shows minimal correlation with hours worked

Number of children has negligible correlation with wages

-   **Gender Differences:**

The education-wage relationship is stronger for males than females

At higher education levels, the gender wage gap appears more pronounced

## [Question 2 = Probability, probability distributions and confidence intervals]{.underline}

#### 2.1

```{r echo=FALSE}

# Randomly select 5 individuals
sampled_individuals <- data[sample(nrow(data), 5), ]

# Display the selected individuals
kable(sampled_individuals)

# Number of simulations
n_simulations <- 10000

# Initialize counter
count_at_least_one_not_covered <- 0

# Run simulation
for (i in 1:n_simulations) {
  sampled <- data[sample(nrow(data), 5), ]
  if (any(sampled$insure == 0)) {
    count_at_least_one_not_covered <- count_at_least_one_not_covered + 1
  }
}

# Estimate the probability
estimated_probability <- count_at_least_one_not_covered / n_simulations

# Display the result
cat("Estimated probability that at least 1 out of 5 is not covered:", round(estimated_probability, 4), "\n")

```

#### 2.2

```{r echo=FALSE}
# Filter to only married individuals (assuming 1 = married)
married <- subset(data, marital == 1)

# Among married, count how many have 1 or more children
num_with_kids <- sum(married$nchild >= 1)

# Total number of married individuals
total_married <- nrow(married)

# Calculate conditional probability
prob_children_given_married <- num_with_kids / total_married

# Display the result
cat("P(1 or more children | married) =", round(prob_children_given_married, 4), "\n")
```

#### 2.3

```{r echo=FALSE}

# Get frequency table for nchild
nchild_table <- table(data$nchild)

# Convert frequency table to probability distribution
nchild_prob <- prop.table(nchild_table)

# Display probability distribution as a table
prob_dist <- data.frame(
  nchild = as.numeric(names(nchild_prob)),
  Probability = as.vector(nchild_prob)
)
print("Probability distribution of nchild:")
kable(prob_dist)

# Calculate mean and variance
mean_nchild <- mean(data$nchild)
var_nchild <- var(data$nchild)

cat("\nMean of nchild:", round(mean_nchild, 4), "\n")
cat("Variance of nchild:", round(var_nchild, 4), "\n")

# Probability that nchild >= 3
p_nchild_ge_3 <- sum(nchild_prob[as.numeric(names(nchild_prob)) >= 3])
cat("P(nchild >= 3):", round(p_nchild_ge_3, 4), "\n")


```

## [Question 3 = Point Estimates, Confidence Intervals & Hypothesis Tests]{.underline}

#### 3.1

```{r echo=FALSE}
# Load the dataset
data <- read.csv("MA334-SP-7_2411779.csv")

# Filter data for households with exactly 2 children
data_2kids <- subset(data, nchild == 2)

# Calculate point estimate (sample mean)
mean_wage <- mean(data_2kids$wage)

# Calculate standard error
se_wage <- sd(data_2kids$wage) / sqrt(nrow(data_2kids))

# 95% confidence interval using normal approximation (or t-distribution for small samples)
conf_int <- mean_wage + c(-1, 1) * qt(0.975, df = nrow(data_2kids) - 1) * se_wage

# Print results
cat("Point estimate (mean wage) for households with 2 children:", round(mean_wage, 2), "\n")
cat("95% Confidence Interval: [", round(conf_int[1], 2), ",", round(conf_int[2], 2), "]\n")

```

#### 3.2

```{r echo=FALSE}
# Load necessary library
library(dplyr)

# Filter households with 5 or more children
large_families <- data %>%
  filter(nchild >= 5)

# View summary or first few rows
kable(large_families)


```

There are only 12 family have 5+ children = **8**

Wages for these households seem to vary widely, from lower wages **(8.25)** to moderate wages **(38.45/hr)**

Gender: **6** male-headed households (gender = 0), **2** female-headed households (gender = 1)

Race: **5** White , **3** Black ,**0** Asian

Work Hours (hrswork) = Most work **40–50** hours/week.

Union Membership (union) = Only **1** household has a union member.

Health Insurance (insure) = **4** out of **8** have health insurance.

#### 3.3

```{r echo=FALSE}


# View unique values for labels
unique(data$insure)
unique(data$gender)

# Create readable labels
data$gender <- factor(data$gender, levels = c(0, 1), labels = c("Male", "Female"))
data$insure <- factor(data$insure, levels = c(0, 1), labels = c("No Insurance", "Has Insurance"))

# Create contingency table
insurance_by_gender <- table(data$insure, data$gender)
colnames(insurance_by_gender) <- c("Male", "Female")
rownames(insurance_by_gender) <- c("No Insurance", "Has Insurance")

# Display table
kable(insurance_by_gender)

# Perform Chi-Squared Test
test_result <- chisq.test(insurance_by_gender)
print(test_result)

test_result$p.value

```

***(Note*** : Gender is coded as **0=Male, 1=Female**; Insurance is coded as **1=Insured, 0=Uninsured)**

Conclusion: Gender and insurance status are not independent (p \< 0.05).

Practical Implication: Males in this dataset are significantly more likely to be insured than females.

data: insurance_by_gender X-squared = **0.80022**, df = **1**, p-value = **0.371**

Test result == **0.3710279**

Interpretation: There is statistically significant evidence at the 5% level that insurance coverage depends on gender.

## [Question 4 = Simple Linear Regression]{.underline}

#### 4.1

```{r echo=FALSE}

# Filter out rows with missing wage or age values
data <- subset(data, !is.na(wage) & !is.na(age) & wage > 0)

# Create groups
young <- subset(data, age < 35)
old <- subset(data, age >= 35)
model_young <- lm(log(wage) ~ age, data = young) # Linear regression for young group
model_old <- lm(log(wage) ~ age, data = old) # Linear regression for old group

# summary
summary(model_young)
summary(model_old)



```

#### 4.2

```{r echo=FALSE , fig.width=3, fig.height=3}


# Remove rows with missing or zero wage
data <- subset(data, !is.na(wage) & wage > 0 & !is.na(age))

# Create log(wage) variable
data$log_wage <- log(data$wage)

# Split into 'young' and 'old'
young <- subset(data, age < 35)
old <- subset(data, age >= 35)

# Fit linear models
model_young <- lm(log_wage ~ age, data = young)
model_old <- lm(log_wage ~ age, data = old)

# Plot: Young group
plot(young$age, young$log_wage, main = "Young (Age < 35): log(Wage) vs Age",
     xlab = "Age", ylab = "log(Wage)", pch = 19, col = "skyblue")
abline(model_young, col = "darkblue", lwd = 2)

# Plot: Old group
plot(old$age, old$log_wage, main = "Old (Age >= 35): log(Wage) vs Age",
     xlab = "Age", ylab = "log(Wage)", pch = 19, col = "salmon")
abline(model_old, col = "darkred", lwd = 2)

# R² values
summary(model_young)$r.squared
summary(model_old)$r.squared

```

## [Question 5 = Multiple Linear Regression]{.underline}

#### 5.1

```{r echo=FALSE}

# Convert categorical variables to factors
data$gender <- factor(data$gender, labels = c("Male", "Female"))
data$race <- factor(data$race)
data$region <- factor(data$region)
data$marital <- factor(data$marital, labels = c("Single", "Married","Other"))
data$insure <- factor(data$insure, labels = c("Uninsured", "Insured"))
data$union <- factor(data$union, labels = c("Non-Union", "Union"))
data$metro <- factor(data$metro, labels = c("Non-Metro", "Metro"))

# Subset the data
young_data <- subset(data, age < 40)
old_data <- subset(data, age >= 40)

# Fit models for log(wage) with all predictors except wage
lm_young <- lm(log(wage) ~ . - wage, data = young_data)
lm_old <- lm(log(wage) ~ . - wage, data = old_data)

# Store summaries
summary_young <- summary(lm_young)
summary_old <- summary(lm_old)

kable(coef(summary_young), digits = 4)

kable(coef(summary_old), digits = 4)
```


We developed distinct multiple linear regression models for each subgroup "young" and "old" to study the relationship between log-transformed wages and various predictors. The age division separates people younger than 40 as "young" and assigns everyone aged 40 or older to "old".

#### 5.2

```{r echo=FALSE}

data <- subset(data, !is.na(wage) & wage > 0)
data$log_wage <- log(data$wage)

# Factor categorical variables
categorical_vars <- c("gender", "educ", "insure", "married", "race")
categorical_vars <- categorical_vars[categorical_vars %in% names(data)]
data[categorical_vars] <- lapply(data[categorical_vars], factor)

# Split into young and old
young <- subset(data, age < 35)
old <- subset(data, age >= 35)

# Full models
full_model_young <- lm(log_wage ~ . - wage - log_wage, data = young)
full_model_old <- lm(log_wage ~ . - wage - log_wage, data = old)

# Stepwise model reduction using AIC
reduced_model_young <- step(full_model_young, direction = "both", trace = FALSE)
reduced_model_old <- step(full_model_old, direction = "both", trace = FALSE)

# Summarize reduced models
kable(coef(reduced_model_young))


kable(coef(reduced_model_old))

```

#### 5.3

Simplicity and Interpretability: Reduced models provide better interpretability which proves especially beneficial when explaining results to audiences without technical backgrounds. The influence of each predictor becomes more apparent when the model contains fewer predictors.

Improved Prediction Accuracy: Eliminating irrelevant or weak predictors from a model can lead to superior performance on new data and enhance its predictive power.

Model Efficiency: Reduced models demand less computational power and exhibit greater numerical stability when working with smaller datasets.


